PoPoolation Pipeline

### 2023-03-21 DRAFT_DNA_methylation.Rmd

---
title: 'DNA Methylation Analyses'
subtitle: '*Steven Roberts*, *Sam White*, and ............ '
bibliography: DNA-Methylation.bib
---

@@

The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
knitr::include_graphics("images/meth-bp-chart.png")
```

For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read. [RRBS data often needs different trimming specifications](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits).

@@

You should expect to prepared genome with new directories including

@@

For aligning the trimmed reads to the genome the following code structure is used

@@

Files are then deduplicated if they are whole genome bisulfite treated samples. This command will deduplicate the Bismark alignment BAM file and remove all reads that align to the the very same position and in the same orientation except for one. This step is recommended for whole-genome bisulfite samples, but **should not** be used for reduced representation libraries such as RRBS, amplicon, or target enrichment libraries. The [`bismark` manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits) provides additional information.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format). The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined. Later in this workflow, a bedGraph with *merged* information will be created.

@@

A genome-wide cytosine report is generated (including merged bedGraph file), with this code as an example:

@@

Generating a file `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)

@@

From here it could be useful to do some simple file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg bedtools, GLMs).

Creating bedGraphs post-merge at a specific coverage threshold (ex. 10x or 5x):

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

@@

Creating tab files with raw count data for GLMs at 10x and 5x coverage:

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_10x.tab
    done

>>>>

---
title: 'DNA Methylation Analyses'
subtitle: '*Steven Roberts*, *Sam White*, and ............ '
bibliography: DRAFT_DNA_methylation/DNA-methylation.bib
---

@@

The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level. The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines. For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
knitr::include_graphics("DRAFT_DNA_methylation/meth-bp-chart.png")
```

@@

You should expect to prepared genome with new directories indluding

@@

For doing the alignment of the trim reads to the genome the following code structured is used

@@

Files are then deduplicated if they are whole genome bisulfite treated samples. This command will deduplicate the Bismark alignment BAM file and remove all reads but one which align to the the very same position and in the same orientation. This step is recommended for whole-genome bisulfite samples, but should not be used for reduced representation libraries such as RRBS, amplicon or target enrichment libraries.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format). And while a bedGraph file is generated, in this workflow a bedGraph will be generated later that uses information from merged strands (as opposed to this output).

@@

A genome-wide cytosine report is generated, with this code as an example:

@@

Generating a file `.CpG_report.merged_CpG_evidence.cov`

@@

From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for downstream analysis (eg bedtools, GLMs).

    #creating bedgraphs post merge
    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

@@

    #creating tab files with raw count for glms
    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_10x.tab
    done

### 2023-03-21 POP_02_RADseq.Rmd

```{r,message = FALSE,echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```
---
title: 'RADseq'
subtitle: '*Katherine Silliman, Danielle Davenport*'
bibliography: RCN_site-Nov05.bib
---

@@

![Table 1 from [@OLeary2018-iy].](images/oleary_table1.png)

@@

Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)

@@

First, we need to convert our VCF file into the GDS format. We will do this once here, then use the GDS file for subsequent code examples.
```{r}
filename = "OL_subset" #replace with your file name 
filename.gds = paste0("RAD_data/", paste0(filename, ".gds"))
filename.vcf = paste0("RAD_data/", paste0(filename, ".vcf"))
  # 1 . Convert VCF to GDS
SeqArray::seqVCF2GDS(vcf.fn = filename.vcf, out.fn = filename.gds, storage.option="ZIP_RA")
gdsin = SeqArray::seqOpen(filename.gds)
print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
```
It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc. In our example, our metadata file (OL.popmap) is tab-delimited and has the column headers:  
ID: sample ID
STRATA: sampling location/population
PLATE: sequencing batch  
Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
```{r}
metafile = "RAD_data/OL.popmap"
sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table(metafile, header = T, sep = "\t") %>%
                  dplyr::select(ID, STRATA, PLATE)
```
Now, on to evaluating our data!

@@

**Missingness in R with SeqArray**
```{r}
#using previously loaded gdsin object
print("Per variant: ")
summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
print("Per sample: ")
summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
samples <- SeqArray::seqGetData(gdsin, "sample.id")
cbind(samples,m2)[order(-m2),]
#plot histogram 
hist(m2,breaks=50)
```

@@

**Missingness with vcftools, on the command line**  
```{bash, eval = T}
vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
# sort the file by most missing data and print the top 10 samples  
cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
```
You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.

To plot the missingness on the command line, you can use gnuplot:
```{bash, eval = T}
#code from Jon Puritz
mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```

@@

```{r}
#open the gds file previously created, if not already open
#filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
#gdsin = SeqArray::seqOpen(filename.gds)
# exclude samples previously identified as having too much missing data
bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

@@

# convert to vcf
seqGDS2VCF(gdsin, "RAD_data/OL-miss-pc1-linkage-filt.vcf.gz")

>>>>

```{r,message = FALSE,echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```
---
title: 'RADseq'
subtitle: '*Katherine Silliman, Danielle Davenport*'
bibliography: POP_02_RADseq_files/common-bib_01.bib
---

@@

![Table 1 from [@OLeary2018-iy].](POP_02_RADseq_files/oleary_table1.png)

@@

Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](POP_02_RADseq_files/lacavaTable3.png)

@@

First, we need to convert our VCF file into the GDS format. We will do this once here, then use the GDS file for subsequent code examples.
```{r}
filename = "OL_subset" #replace with y our file name 
filename.gds = paste0("POP_02_RADseq_files/", paste0(filename, ".gds"))
filename.vcf = paste0("POP_02_RADseq_files/", paste0(filename, ".vcf"))
  # 1 . Convert VCF to GDS
SeqArray::seqVCF2GDS(vcf.fn = filename.vcf, out.fn = filename.gds, storage.option="ZIP_RA")
gdsin = SeqArray::seqOpen(filename.gds)
print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
```
It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc. In our example, our metadata file (OL.popmap) is tab-delimited and has the column headers:  
ID: sample ID
STRATA: sampling location/population
PLATE: sequencing batch  
Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
```{r, eval = T}
metafile = "POP_02_RADseq_files/OL.popmap"
sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table(metafile, header = T, sep = "\t") %>%
                  dplyr::select(ID, STRATA, PLATE)
```
Now, on to evaluating our data!

@@

**Missingness in R with SeqArray**
```{r, eval = T}
#using previously loaded gdsin object
print("Per variant: ")
summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
print("Per sample: ")
summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
samples <- SeqArray::seqGetData(gdsin, "sample.id")
cbind(samples,m2)[order(-m2),]
#plot histogram 
hist(m2,breaks=50)
```

@@

**Missingness with vcftools, on the command line**  
```{bash, eval = T}
vcftools --vcf POP_02_RADseq_files/OL_subset.vcf --missing-indv --out POP_02_RADseq_files/OL_subset
# sort the file by most missing data and print the top 10 samples  
cat POP_02_RADseq_files/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
```
You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.

To plot the missingness on the command line, you can use gnuplot:
```{bash, eval = T}
#code from Jon Puritz
mawk '!/IN/' POP_02_RADseq_files/OL_subset.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```

@@

```{r}
#open the gds file previously created, if not already open
#filename.gds = "POP_02_RADseq_files/OL_subset.gds" #replace with your file name 
#gdsin = SeqArray::seqOpen(filename.gds)
# exclude samples previously identified as having too much missing data
bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
sample.ids = seqGetData(gdsin, "sample.id")
keep = sample.ids[which(!sample.ids %in% bad_samples)]

@@

# convert to vcf
seqGDS2VCF(gdsin, "POP_02_RADseq_files/OL-miss-pc1-linkage-filt.vcf.gz")

### 2023-03-21 POP_03_poolseq.Rmd

---
title: "PoPoolation Pipeline"
subtitle: '*Joanna Griffiths*'
date: "8/12/2021"
bibliography: RCN_site-Nov05.bib
---

@@

```
![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](images/Fst_SvsBR_SDmap.png)

@@

```
![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](images/pi_PCA.png)

>>>>

---
title: "PoPoolation Pipeline"
subtitle: '*Joanna Griffiths*'
date: "8/12/2021"
bibliography: "POP_03_poolseq_files/common-bib_01.bib"
---

@@

```
![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](POP_03_poolseq_files/Fst_SvsBR_SDmap.png)

@@

```
![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](POP_03_poolseq_files/pi_PCA.png)
